---
title       : Human Activity Recognition - Exploring the Dataset
subtitle    : Developing Data Projects Course
author      : Andrey Boytsov
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

```{r echo=FALSE, results='hide'}
## ====== READING AND CLEANING THE DATA =====
xLabeled <- read.csv("pml-training.csv")
beforeCleaning <- ncol(xLabeled)-1

# Let's make it verbose for better understanding

# 1. X column is just a sequential number. Do not use.
xLabeled <- xLabeled[, !(colnames(xLabeled)=="X")]
N = length(xLabeled[,1])

#2. Timestamp columns should not be used
columnsToDrop <-c("raw_timestamp_part_1","raw_timestamp_part_2","cvtd_timestamp","X","user_name","new_window", "num_window");
xLabeled <- xLabeled[, !(colnames(xLabeled) %in% columnsToDrop)]

beforeSensorCleaning <- ncol(xLabeled)-1

#In the report - use first few lines of dataset. Show that we have NAs and empty values.

#3. Remove columns with too many missing values
# In the report - show table with percentage (separate of joint), it will provide good illustration
cutoff_percentage = 0.95 #Let's say, over 95% of NA or empty values means that column should be dropped
# In the report - use "at least 1" cutoff
cutoff_percentage <- 0.95
xLabeled <- xLabeled[, lapply(xLabeled, function(x) (sum(is.na(x) | x=="")) / length(x)) < cutoff_percentage]
xLabeled <- xLabeled[complete.cases(xLabeled), ]

#4. Generate activities list
activitiesListForCheckbox <- as.list(as.character(unique(xLabeled$classe)))
activityFriendlyNames <- c("Sitting down (class A)","Standing up (class B)", "Standing (class C)","Walking (class D)","Sitting (class E)");

#5. Generate selection parameters out of column names
selectorParameters <- colnames(xLabeled);
selectorParameters <- selectorParameters[1:(length(selectorParameters)-1)]
selectorParametersFriendlyNames <- selectorParameters
toReplace <- c("yaw_", "pitch_", "roll_", "_x", "_y", "_z", "total_accel_", "accel_", "magnet_", "gyros_")
replacement <- c("Yaw - ", "Pitch - ", "Roll - ", " - X-axis", " - Y-axis", " - Z-axis", "Absolute acceleration - ", "Accelerometer - ", "Magnetometer - ", "Gyroscope - ")
for (i in 1:length(selectorParametersFriendlyNames)){
  for (j in 1:length(toReplace)){
      selectorParametersFriendlyNames[i] <- gsub(toReplace[j], replacement[j], selectorParametersFriendlyNames[i])
  }
}
names(selectorParameters) <- selectorParametersFriendlyNames;
```

## Basic Information

 * Purpose
  * Get ready for Practical Machine Learning course
   * Understand the dataset
   * Explore it
   * Or review the lessons, if you already took that course
  * Explore real-life human activity recognition dataset
   * Find the sensed features that can distinguish one activity from another
 * Dataset
  * Human Activity Recognition Dataset collected by Pontifical Catholic University of Rio de Janeiro
(Rio de Janeiro, Brazil) - http://www.puc-rio.br/
  * Dataset reference: Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

--- .class #id 

## Dataset Details

 * Dataset available under Creative Commons license (CC-BY-SA)
 * Contains information about following excercises: `r activityFriendlyNames`
 
```{r echo=FALSE, results='asis'}
par(mar = c(7, 12, 1, 0));
barplot(table(xLabeled$classe), xlab = "Excercise Type", main = "Excercise Type Count", col = "red", horiz = T, names.arg = activityFriendlyNames, las=1)
```

--- .class #id 

## Dataset and Applicaiton

 * Sensor positions: arm, belt, forearm, dumbbelt
 * Sensor types: accelerometer, magnetometer, gyroscope
 * Dataset contains following potential information sources (sensor readings): `r selectorParametersFriendlyNames[1:10]`, etc.
   * and many more...
 * Application already performed initial data cleaning
  * Before any data cleaning: `r beforeCleaning` available information sources
  * Before analyzing sensor reliability: `r beforeSensorCleaning` available information sources
  * After full data cleaning: `r length(selectorParametersFriendlyNames)` available sensors
 * Application transformed names to friendly readable format
  * Both for sensor names and for activity names
  * Transformation is automated

--- .class #id 

## Using The Application
  
  * Select the activities of interest using the checkboxes
    * The activities selected on the left will be depicted in red
    * The activities selected on the right will be depicted in blue
  * Select the sensor reading in the drop-down menu
    * Sensor readings that have too many N/As are already removed from the list
    * Unrelated information sources (like number of the sliding window) are also removed from the list
  * Analyze the histogram
    * Select at least one activity for histogram to appear
    * Turn on mean or median lines if required
  * Analyze summary statistics at the bottom


